A Comparative Quality Metric for Fuzzing with Logic State Coverage

### 2024-03-03 paper/main.tex
interestingness: removed 1/n remark
paper: revised interestingness

Fuzzing, as a form of software testing, aims at minimizing the number of unknown
abnormal program behaviors. \emph{Untargeted} fuzzing addresses this goal by:
(i) investigating as many \emph{new} program behaviors as possible and (ii)
keeping the new behaviors \emph{more interesting} (i.e., more likely revealing
unknown abnormal behaviors).

\pp{Formal justification}
%
The approach can be described formally as follows. Let the
\emph{interestingness} of a program behavior be $P(\mathrm{abnormal})$, the
expectation of a certain program behavior being abnormal before observing
it.\footnote{See \autoref{s:disc} for the compatibility of this definition and
the conventional meaning of interestingness in untargeted fuzzing.} Furthermore,
let $P(\mathrm{new})$ be the probability of a new behavior in the next
execution.  Then, the probability of discovering unknown abnormal behavior in
the next execution is
%
\begin{equation}
  P(\mathrm{new} \cap \mathrm{abnormal}) = P(\mathrm{new}) \times P(\mathrm{abnormal} \mid \mathrm{new}).
\end{equation}

In this formulation, untargeted fuzzing attempts to (i) increase
$P(\mathrm{new})$ to investigate more program behaviors while (ii) keep 
$P(\mathrm{abnormal} \mid \mathrm{new})$ as high as possible to ensure new
behaviors interesting.
%
Notice that executing more inputs is not the same as observing more distinct
behaviors, as inputs can exhibit the same behavior (e.g., the same
early-termination).

\pp{Interestingness in practice}
%
As the interestingness $P(\mathrm{abnormal})$ depends on the expectation of an
\emph{unobserved} behavior, there is no direct way to measure or formalize
interestingness. However, the practice of untargeted fuzzing in recent years
suggests that abnormal behaviors get increasingly unlikely to require a
larger number of branch hit.
%
The prime example is the branch hit count bucketing in AFL \cite{afl}, where the
bucket size gets larger roughly by the power of 2 (i.e., 1, 2, 3, 4-7, 8-15,
...).
%; all hit counts are treated the same within a bucket.
%
%This is also intuitive because if a certain abnormal behavior can manifest at a
%low loop count, such a behavior is not \emph{unknown} anymore by when it reaches
%a higher loop count.

To formalize this, let $X=n$ denote when a program behavior takes a 
branch $X$ $n$ times ($n \ge 1$). Then, the interestingness of a program
behavior when $X=n$ is;
%
\begin{equation}
  P(\mathrm{abnormal} \mid X=n) \propto 1/n.
\end{equation}

>>>>

Fuzzing, as a form of software testing, aims at minimizing the number of unknown
abnormal program behaviors. \emph{Untargeted} fuzzing addresses this goal by:
(i) investigating as many \emph{new} program behaviors as possible and (ii)
keeping the mutation base input \emph{interesting} (i.e., more likely revealing
unknown abnormal behaviors after mutation).

\pp{New vs. unknown behaviors}
%
We first make a clear distinction between \emph{new} and \emph{unknown} 
behaviors. \emph{New} behaviors are, as the name suggests, simply the behaviors 
that have not been observed until now. On the other hand, \emph{unknown}
behaviors are a subset of the \emph{new} behaviors \emph{not equivalent} to any
observed behaviors. In particular, an abnormal behavior is \emph{unknown} if it
is not equivalent to any observed abnormal behaviors (e.g., a unique root
cause). Formally, if $\mathrm{U}$ and $\mathrm{N}$ are the sets of
\underline{\textbf{u}}nknown and \underline{\textbf{n}}ew behaviors,
respectively, $\mathrm{U} \subset \mathrm{N}$.

\pp{Formal justification}
%
The approach of untargeted fuzzing can be described formally as follows. Upon
executing a mutated input, let $P(\mathrm{N})$ be the probability of observing a
\underline{\textbf{n}}ew behavior as a result of execution. Then, the
probability of discovering \underline{\textbf{u}}nknown
\underline{\textbf{a}}bnormal behavior $P(\mathrm{UA}) := P(\mathrm{U} \cap
\mathrm{A})$ is
%
\begin{align}
  P(\mathrm{UA}) 
      &= P(\mathrm{N} \cap \mathrm{UA}) \tag{by $\mathrm{U} \subset \mathrm{N}$} \\
      &= P(\mathrm{N}) \times P(\mathrm{UA} \mid \mathrm{N}) \tag{by chain rule}. 
\end{align}

Here, the second term $P(\mathrm{UA} \mid \mathrm{N})$ is the expectation of
revealing an unknown abnormal behavior given a new behavior, which can be
interpreted as the \emph{interestingness} of the mutation base. In other words,
untargeted fuzzing attempts to (i) increase $P(\mathrm{N})$ to investigate more
program behaviors while (ii) keeping the interestingness $P(\mathrm{UA} \mid
\mathrm{N})$ as high as possible.
%
%Notice that executing more inputs is not the same as observing more distinct
%behaviors, as inputs can exhibit the same behavior (e.g., the same
%early-termination).

\pp{Interestingness in practice}
%
As the interestingness $P(\mathrm{UA} \mid \mathrm{N})$ is the expectation of an
\emph{unobserved} behavior, there is no direct way to measure interestingness. 
%
However, the practice of untargeted fuzzing suggests that
the resulting program behavior is unlikely to be \emph{unknown abnormal} if the
mutation base is highly repetitive, as it commonly incorporates corpus
minimization \cite{} and branch hit count buckets \cite{}.
%
The rationales are: (i) abnormal behaviors might have been already observed with
fewer branch repetitions (i.e., unlikely \emph{abnormal}) and (ii) if so, it may
correspond to pre-observed less-repetition counterparts (i.e., unlikely
\emph{unknown}).

%The prime example is the branch hit count bucketing in AFL \cite{afl}, where
%the bucket size gets larger roughly by the power of 2 (i.e., 1, 2, 3, 4-7,
%8-15, ...).

%However, the practice of untargeted fuzzing in recent years suggests that
%abnormal behaviors get increasingly unlikely to require a larger number of
%branch hit.
%
%The prime example is the branch hit count bucketing in AFL \cite{afl}, where
%the bucket size gets larger roughly by the power of 2 (i.e., 1, 2, 3, 4-7,
%8-15, ...).
%; all hit counts are treated the same within a bucket.
%
%This is also intuitive because if a certain abnormal behavior can manifest at a
%low loop count, such a behavior is not \emph{unknown} anymore by when it reaches
%a higher loop count.

%To formalize this, let $X=n$ denote when a program behavior takes a 
%branch $X$ $n$ times ($n \ge 1$). Then, the interestingness of a program
%behavior when $X=n$ is;
%%
%\begin{equation}
%  P(\mathrm{abnormal} \mid X=n) \propto 1/n.
%\end{equation}
